---
title: "STA 313 - Spring 2026 - HW 5"
subtitle: "Mirror mirror on the wall..."
author: "Alain Fornes"
format: html
engine: knitr
toc: true
---

## Part 1 - Who’s the happiest of them all?

### Question 1 

**The data.**

```{r}
#| label: include-data-prep-r-script
#| eval: false
#| file: data/data-prep.R
```

### Question 2

**The visualization.** 

```{r}
#| label: load-packages-r
#| message: false

library(tidyverse)
library(ggthemes)
```

```{r}
#| label: load-data-r
#| message: false

whr <- read_csv("data/whr.csv")
```

```{r}
#| label: q2-1
#| message: false

whr_2025 <- whr |>
  filter(year == 2025)

leader_countries <- bind_rows(
  whr_2025 |> slice_max(explained_by_log_gdp_per_capita, n = 1),
  whr_2025 |> slice_max(explained_by_social_support, n = 1),
  whr_2025 |> slice_max(explained_by_healthy_life_expectancy, n = 1),
  whr_2025 |> slice_max(explained_by_freedom_to_make_life_choices, n = 1),
  whr_2025 |> slice_max(explained_by_generosity, n = 1),
  whr_2025 |> slice_max(explained_by_perceptions_of_corruption, n = 1),
  whr_2025 |> slice_max(dystopia_residual, n = 1)
) |>
  distinct(country, .keep_all = TRUE) |>
  mutate(
    country = fct_reorder(country, life_eval)
  )

leaders_long <- leader_countries |>
  select(
    country,
    life_eval,
    explained_by_log_gdp_per_capita,
    explained_by_social_support,
    explained_by_healthy_life_expectancy,
    explained_by_freedom_to_make_life_choices,
    explained_by_generosity,
    explained_by_perceptions_of_corruption,
    dystopia_residual
  ) |>
  pivot_longer(
    cols = explained_by_log_gdp_per_capita:dystopia_residual,
    names_to = "component",
    values_to = "value"
  ) |>
  mutate(
    component = recode(
      component,
      explained_by_log_gdp_per_capita = "Log GDP per capita",
      explained_by_social_support = "Social support",
      explained_by_healthy_life_expectancy = "Healthy life expectancy",
      explained_by_freedom_to_make_life_choices = "Freedom to make life choices",
      explained_by_generosity = "Generosity",
      explained_by_perceptions_of_corruption = "Perceptions of corruption",
      dystopia_residual = "Dystopia + residual"
    )
  )

ggplot(leaders_long, aes(x = value, y = country, fill = component)) +
  geom_col() +
  scale_x_continuous(
    limits = c(-0.3, NA)
  ) +
  scale_fill_colorblind() +
  labs(
    title = "How happiness is composed for countries leading in key happiness components in the year 2025",
    subtitle = "Each country leads overall in one explanatory compnent of happiness for the year 2025",
    x = "Contribution to life evaluation score",
    y = NULL,
    fill = "Component",
    caption = "Source: World Happiness Report 2026"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

The visualization I made zooms ina nd focuses on the countries that lead at least one key explanatory component of happiness in 2025. This visualization, similar to the original, also shows how each country’s life evaluation score is built from those different pieces. It reveals that these countries do not all achieve high happiness in the same way, since some stand out more in areas some components than others, and the visualization helps us see how that contribution is distributed in thse different leaders.

**AI reference disclosure**
I asked it to help write code to answer the question.
https://chatgpt.com/share/69e65ddf-abf4-83ea-baa6-272523a5d30d

**Non-AI reference disclosure**
Figure 2.1 from the report.

### Question 3

**The app.** 

**App Link**
https://connect.posit.cloud/alain-fornes/content/019dac51-4b0c-2c68-f08d-8ea3b3806a24

```{r}
#| label: include-app-r-script
#| eval: false
#| file: app/app.R
```

**AI reference disclosure**
I asked it to help write code to answer the question.
https://chatgpt.com/share/69e67840-1314-83ea-9c69-8b27e699aeb6

**Non-AI reference disclosure**
https://docs.posit.co/connect-cloud/user/account/github.html

## Part 2 - Who’s the ugliest of them all?

### Question 4

**Ugly, but in R.** 

```{r}
#| label: ugly-plot-r
three_countries <- whr |>
  filter(country %in% c("Finland", "United States", "Afghanistan"))

ggplot(three_countries, aes(x = year, y = life_eval, color = country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = unique(three_countries$year)) +
  labs(
    title = "Life evaluation over time for three countries",
    x = "Year",
    y = "Life evaluation score",
    color = "Country"
  ) +
  theme_minimal()
```

```{r}
#| label: q4-ugly-plot

ggplot(three_countries, aes(x = year, y = life_eval, color = country)) +
  geom_line() +
  geom_point(aes(shape = country), size = 5) +
  scale_color_manual(values = c("red", "limegreen", "purple")) +
  labs(
    title = "HAPPINESS GRAPH!",
    subtitle = "COMPARISON of HapPiness OVER TIME!!!!!!!!!",
    x = "YEAR OF OBSERVATION (missing 2013 for som reason)",
    y = "LIFE EVALUATION (3-YEAR AVERAGE) SCORING!!!!",
    color = "WHICH COUNTRY??",
    shape = "COUNTRY AGAIN!!!"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 22, face = "bold.italic", hjust = 0.5, color = "blue"),
    plot.subtitle = element_text(size = 14, face = "italic", color = "darkorange"),
    axis.title = element_text(size = 16, face = "bold", color = "darkred"),
    axis.text.x = element_text(angle = 60, hjust = 1, size = 12, color = "magenta"),
    axis.text.y = element_text(size = 12, color = "forestgreen"),
    legend.position = "bottom",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    panel.background = element_rect(fill = "lightyellow"),
    plot.background = element_rect(fill = "lightpink"),
    panel.grid.major = element_line(color = "cyan", linewidth = 1),
    panel.grid.minor = element_line(color = "orange", linewidth = 0.8)
  )
```

**AI reference disclosure**
I asked it to help write code to answer the question.
https://chatgpt.com/share/69e67e5c-2090-83ea-ba1d-56e851c0a307

**Non-AI reference disclosure**
I did not use a Non-AI reference to answer this question.

### Question 5

**Ugly, but in Python.** 

```{r}
#| label: py-setup
#| include: false

library(reticulate)

py <- normalizePath("./.venv/bin/python", mustWork = TRUE)
Sys.setenv(RETICULATE_PYTHON = py)
use_python(py, required = TRUE)
py_config()
```

```{python}
#| label: load-packages-py
#| message: false

import pandas as pd
import plotnine as p9
```

```{python}
#| label: load-data-py
#| message: false

whr = pd.read_csv("data/whr.csv")

three_countries = whr[whr["country"].isin(["Finland", "United States", "Afghanistan"])].copy()
year_breaks = sorted(three_countries["year"].unique())
```


```{python}
#| label: pretty-plot-py

(
    p9.ggplot(three_countries, p9.aes(x="year", y="life_eval", color="country"))
    + p9.geom_line()
    + p9.geom_point()
    + p9.scale_x_continuous(breaks=year_breaks)
    + p9.labs(
        title="Life evaluation over time for three countries",
        x="Year",
        y="Life evaluation score",
        color="Country"
    )
    + p9.theme_minimal()
    + p9.theme(
        figure_size=(8, 5),
        axis_text_x=p9.element_text(size=8)
    )
)
```

```{python}
#| label: ugly-plot-py

(
    p9.ggplot(three_countries, p9.aes(x="year", y="life_eval", color="country"))
    + p9.geom_line()
    + p9.geom_point(p9.aes(shape="country"), size=4.5)
    + p9.scale_x_continuous(breaks=year_breaks)
    + p9.scale_color_manual(values=["red", "limegreen", "purple"])
    + p9.labs(
        title="HAPPINESS GRAPH!",
        subtitle="COMPARISON of HapPiness OVER TIME!!!!!!!!!",
        x="YEAR OF OBSERVATION (missing 2013 for som reason)",
        y="LIFE EVALUATION (3-YEAR AVERAGE) SCORING!!!!",
        color="WHICH COUNTRY??",
        shape="COUNTRY AGAIN!!!"
    )
    + p9.theme_bw()
    + p9.theme(
        figure_size=(2, 7),
        plot_title=p9.element_text(
            size=16, weight="bold", style="italic", ha="center", color="blue"
        ),
        plot_subtitle=p9.element_text(
            size=11, style="italic", color="darkorange"
        ),
        axis_title=p9.element_text(
            size=11, weight="bold", color="darkred"
        ),
        axis_text_x=p9.element_text(
            rotation=60, ha="right", size=9, color="magenta"
        ),
        axis_text_y=p9.element_text(
            size=9, color="forestgreen"
        ),
        legend_position="bottom",
        legend_title=p9.element_text(
            size=10, weight="bold"
        ),
        legend_text=p9.element_text(
            size=9
        ),
        panel_background=p9.element_rect(fill="lightyellow"),
        plot_background=p9.element_rect(fill="lightpink"),
        panel_grid_major=p9.element_line(color="cyan", size=0.8),
        panel_grid_minor=p9.element_line(color="orange", size=0.6),
        subplots_adjust={
            "left": 0.22,
            "right": 0.98,
            "bottom": 0.32,
            "top": 0.88
        }
    )
)
```

I was able to almost create full replicas of the plots built in R, save a few issues. First, the quality of the renders is not the same, with some words being a bit more pixelated in the python plots. Also, The legend on the second plot (the ugly one) was impossible to make fit withing the pink box. I tried using many online resources but nothing I found could help me make the legend not span the entirety of the render.

**AI reference disclosure**
I did not use AI references for this part of the assignment

**Non AI reference disclosure**
https://plotnine.org/